### 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/revisions/2022-03/response.tex

%% General paragraph settings (no indent, big line skip).
\parindent=0pt
\parskip=9pt
\baselineskip=13pt plus 1pt minus 1pt \lineskiplimit=1pt

%% Red coloring.
\input colordvi
\def\BEGINC{\textRed}
\def\ENDC{\textBlack}
\def\red#1{\BEGINC#1\ENDC}
\let\beginred=\BEGINC
\let\endred=\ENDC

%% Fonts (for title)
\font\titlebf=cmbx20
\font\titlett=cmtt20
%% Heading font.
\font\headingbf=cmbx14
\font\headingtt=cmtt14
%% Caption font.
\font\caprm=cmr8
\font\captt=cmtt8
\font\capit=cmmi8
\font\capbf=cmbx8
\font\capsy=cmsy8
%% Caption text.
\def\captext{
  \narrower
  \noindent
  \caprm
  \baselineskip=10pt plus 2pt minus 1pt
  \lineskiplimit=2pt
  \lineskip=2pt plus 1pt
  \parskip=0pt
  \textfont0=\caprm
  \textfont1=\capit
  \textfont2=\capsy
}

@@

\itemitem{} Costantini, P. An algorithm for computing shape-preserving
  interpolating splines of arbitrary degree, Journal of Computational
  and Applied Mathematics, 1998, 22(1), pp. 89-136.

\item{} While the concerns about subjectivity in comparisons are
  acknowledged, users will still want to know the context of this work
  and the trade-offs would they use this package instead of other
  spline algorithms. For this reason a comparison with other methods
  including visuals should be provided.

}

We appreciate this perspective and have included such comparisons in
the updated Section 5 (Figures 5-8). The comparison shows methods from
four published spline software packages against MQSI for the same test
problems as originally in the paper (and associated test code). An
important point to note is that we were hesitant to include
comparisons with the published R package for Schumaker spline
interpolants, because (as is shown in the revised manuscript) the
package often produces incorrectly nonmonotone approximations to
data. In our opinion this undesirable outcome is a valid demonstration
of the numerical difficulties that arise when writing codes for spline
algorithms. They are not trivial to handle, and that underscores the
importance of thoroughly and carefully peer reviewed implementations
like what is provided by this work.

@@

\goodbreak
{\parindent=20pt \it

\item{(3)} The monotonicity of Q(x) is not proved.

}

Per Section 1 paragraph 5, Section 2 paragraphs 3, 4 and 6, and
Algorithm 2 (all of the original submission) the theory proving the
sharp necessary and sufficient conditions for monotonicity is provided
by Ulrich and Watson [1994]. This code conforms precisely with the
tight boundary conditions referenced therein. The theory is quite
dense and it would seem inappropriate to reproduce all of that work in
this manuscript as well. To further clarify that the conditions for
monotonicity are proven, we have added an additional paragraph to
Section 4 explaining this fact.

@@

\goodbreak
{\parindent=20pt \it

\item{(4)} The approximation order of the algorithm is not shown.

}

In the same paragraph of Section 4 referenced above, we have explained
that the approximation order of the method is that of any second order
approximation (because a local quadratic interpolant is used to
approximate initial derivative values), but also note that the
asymptotic approximation order is often of little importance in the
context of (sparse) scattered data (the target application of most
shape preserving techniques, especially those achieving higher levels
of continuity).

@@

\goodbreak
{\parindent=20pt \it

\item{(6)} Some statements depend on two variables being
  ``approximately equal'', but what does that mean? Equal up to some
  small threshold?

}

The definition of approximately equal here is when two floating point
numbers are within the machine precision (at their scale) of each other.
This is a rather standard approach for ensuring safe separation of
floating point numbers. We have added a sentence explaining this to
the pretexts of the algorithms that reference this operation.

@@

{\parindent=20pt \it

\item{(9)} Your binary search results in a particular set of
  derivatives, for which the local quintic piece is guaranteed to be
  monotone, but it is clearly not a unique choice. Alternatively, one
  could try to find, among all possible sets of derivatives and
  corresponding monotone quintic interpolants, the best, according to
  some criterion, e.g. least L2 norm. This would turn the problem into
  a constrained optimization problem, which is probably harder to
  solve, but it might, at the same time, give better results and
  overcome the limitation shown in Fig. 1. More motivating details
  would be appreciated.

}

This was a point of our own severe consideration before submitting the
manuscript and this comment reflected our original thoughts. In due
response to this we've implemented a naive $L^2$ minimization algorithm
over the second derivative that uses gradient descent over the spline
representation to enforce local monotonicity constraints in a
Lagrangian fashion (with an adaptive multiplier on monotonicity
violations). The results we achieved were quite mixed and we hesitate
to incorporate commentary from this excursion into the manuscript. In
truth, formally expanding on this trajectory would require an entirely
new manuscript with deeper study of the problem, while it is not
immediately obvious that the results would be {\it better} than the
present approach. The global $L^2$ minimizer of the second derivative
is simply the unique $C^2$ cubic spline interpolant, but that is known
to produce large and unnecessary local variations in the approximation
(along with not being shape preserving). Included in this response (to
reviewers only) are some of the preliminary visuals we produced in
exploring this approach. This is by no means final and conclusive, but
the results were mixed enough for us to warrant not including them in
the paper.

\vfill

\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/5-l2_minimizer.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/5-l2_minimizer.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/5-l2_minimizer.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/5-l2_minimizer.eps}}
{\captext {\capbf $L^2$ Minimization Experiment.} {\captt MQSI} (light
  gray) compared with an experimental global $L^2$ minimization
  approach (blue) on each of the test problems from the paper. It is
  {\capit very} difficult to consistently construct fits that lack
  local oscillations (i.e., minimize $L^2$) as well as maintain
  monotonicity with this approach. The method relies on the B-spline
  basis in this experiment, which affects the bias towards specific
  solutions when following the second derivative $L^2$ minimizing
  gradient. This work would warrant a significant line of research to
  properly resolve these issues in a theoretically sound way.
\par}
\vfill \eject

@@

\goodbreak
{\parindent=20pt \it

\item{(10)} It is quite probable that a user would want to convert the
  piecewise Hermite representation of the interpolant to quintic
  B-spline form. But the authors say in Sections 3 and 4 that they
  solve a banded linear system of length n (n the number of data
  points) to make this conversion. However, this is not necessary. At
  each data point, only three of the quintic B-splines are
  non-zero. Therefore, the three associated B-spline coefficients can
  be computed from the value, first, and second derivatives of the
  interpolant at that single point. Thus one can just solve a small
  3*3 linear system to get one group of 3 B-spline coefficients. Well,
  it requires a bit of algebraic work to write out this 3*3 linear
  system, but it can be done.

}

As mentioned in the manuscript Section 5 paragraph 4, the banded
linear system representation is favored for it's decreased redundancy.
The computational complexity order is unchanged by solving a single
banded system, while the global numerical stability is {\it
  theoretically} improved. The system of equations and number of
nonzero splines at each breakpoint is thoroughly documented in
{\tt SPLINE.f90}, lines 172--206. In practice this approach is
almost certainly slower than solving individual linear systems, but
the increased global stability and decreased redundancy in
representation is favored here (since the banded linear system also
does not increase computational complexity). For users interested in
constructing alternative representations, the optional argument {\tt
  UV} documented in the code exists precisely for returning the
computed first and second derivatives at knots.

@@

\item{(12)} The output are values -- why not the spline coefficients? If
  the output are values one would hope for a visualization package.

}

The code includes suitable documentation and examples to conform to
any external visualization package desired by a user. Given the wide
variety of mechanisms by which a user might want to visualize or use
the results of this code, it is nearly impossible to provide a good
generic solution to visualization.

For the convenience of reviewers, we are providing the code with which
we generated the visuals for this manuscript. Although we note it has
very specific system requirements to operate effectively. Namely: (1)
running a POSIX shell (on macOS, Ubuntu, $\ldots$); (2) having {\tt
  python3} of at least version $3.7$ and the packages {\tt numpy},
{\tt tlux}, and {\tt rpy2}; (3) having installed {\tt gfortran} (or
otherwise reconfiguring the codes to use another Fortran compiler of
choice); and (4) having installed {\tt R} (for Schumaker splines). The
Fortran codes associated with this work as well as those from TOMS 574
(quadratic $C^1$) and BVSPIS from TOMS 770 are all wrapped to be
callable directly from {\tt python3}, which will be the parent process
for all calls. A simple test usage can be seen in the provided {\tt
  python3} file {\tt test.py}. To be abundantly clear, this code is
{\bf not} intended for review and publication, but rather convenient
access to various codes for reviewers to use for visualizing and
comparing results. Precise installation command line instructions are
provided at the end of this document.

@@

}

The term ``univariate'' has been added to the abstract for greater
clarity on this point.

\vfill \eject

Throughout the manuscript, all modified text has been marked in
\red{red}. We hope that you all find these revisions to have improved
the manuscript, and we look forward to your response.

@@

Sincerely,

Thomas Lux, on behalf of all authors.

\vfill

{\headingbf Installing dependencies and running tests in} {\hskip 2mm} {\headingtt MQSI\_testing.zip}

\vskip 5mm

{\tt Ubuntu 20+} command line instructions.

{\parskip=2pt \parindent=20pt \tt
\item{} \#() \string{ echo "\# \$@" \string}
\item{} \textMaroon \# install gfortran \textBlack
\item{} sudo apt install gfortran
}

\vskip 5mm

{\tt macOS 12+} command line instructions (installs CommandLineTools and Homebrew first).

{\parskip=2pt \parindent=20pt \tt
\item{} \#() \string{ echo "\# \$@" \string}
\item{} \textMaroon \# install CommandLineTools if you do not have it already \textBlack
\item{} xcode-select -p || xcode-select --install
\item{} \textMaroon \# install Homebrew for package management \textBlack
\item{} brew --version || /bin/bash -c "\$(curl -fsSL $\backslash$
\item{} \hskip 10mm https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
\item{} \textMaroon \# install gfortran \textBlack
\item{} brew install gfortran
}

\vskip 5mm

Common instructions: Python package installation, unzipping archive, and running {\hskip 1mm} {\tt test.py}.

{\parskip=2pt \parindent=20pt \tt
\item{} \textMaroon \# install a visualization python package \textBlack
\item{} python3 -m pip install --user tlux
\item{}
\item{} \textMaroon \# unzip archive, enter resulting directory, run tests \textBlack
\item{} unzip MQSI\_testing.zip
\item{} cd MQSI\_testing
\item{} python3 test.py
\item{}
\item{} \textMaroon \# install R (if comparisons with Schumaker are desired) \textBlack
\item{} \textMaroon \# install Schumaker package through R graphical user interface \textBlack
\item{} python3 -m pip install --user rpy2
\item{}
\item{} \textMaroon \# make any desired modifications to test data at bottom of \textBlack test.py 
}

### 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
All testing code and last comments implemented.

Many domains of science rely on smooth approximations to real-valued
functions over a closed interval. Piecewise polynomial functions
(splines) provide the smooth approximations for animation in graphics
[Herman et al.\ 2015; Quint 2003], aesthetic structural support in
architecture [Brennan 2020], efficient aerodynamic surfaces in
automotive and aerospace engineering [Brennan 2020], prolonged
effective operation of electric motors [Berglund et al.\ 2009], and
accurate nonparametric approximations in statistics [Knott
2012]. While polynomial interpolants and regressors apply broadly,
splines are often a good choice because they can approximate globally
complex functions while minimizing the local complexity of an
approximation.

It is often the case that the true underlying function or phenomenon
being modeled has known properties like convexity, positivity,
various levels of continuity, or monotonicity. Given a reasonable
amount of data, it quickly becomes difficult to achieve desirable
properties in a single polynomial function. In general, the
maintenance of function properties through interpolation/regression is
referred to as {\it shape preserving} [Fritsch and Carlson 1980;
Gregory 1985]. The specific properties the present algorithm will
preserve in approximations are monotonicity and $C^2$ continuity. In
addition to previously mentioned applications, these properties are
crucially important in statistics to the approximation of a cumulative
distribution function and subsequently the effective generation of
random numbers from a specified distribution [Ramsay 1988].  A spline
function with these properties could approximate a cumulative
distribution function to a high level of accuracy with relatively few
intervals. A twice continuously differentiable approximation to a
cumulative distribution function (CDF) would produce a corresponding
probability density function (PDF) that is continuously
differentiable, which is desirable.

The existing research in shape preserving interpolatory splines is
rich and filled with many approaches for many different
applications. In the context of this work, the unique quality of shape
preserving spline algorithms can be observed through: (1) the type of
spline (rational or Hermite), (2) the polynomial order and level of
continuity, and (3) the sharpness, sufficiency, and necessity of the
conditions used to establish shape preserving properties. While the
choice of spline representation is less important, achieving higher
levels of continuity is often more difficult than lower levels of
continuity. And while all referenced algorithms establish sufficient
conditions for shape preservation, it is less common and much more
difficult to provide sharp {\it necessary} conditions for
monotonicity. The following two paragraphs highlight a collection of
research related to the construction of sufficient conditions for
$C^1$ and $C^2$ shape preserving splines. From this foundation, the
value and utility of the code presented here can be more readily
appreciated.

DeVore [1977] proves that monotone spline interpolants have improved
accuracy over their nonmonotone counterparts when approximating
monotone functions, while Constantini [1986] proves that monotone
spline interpolants to data of fixed continuity and order exist in
general. This groundwork establishes the potential for monotone spline
interpolation. Schumaker [1983] and McAllister and Roulier [1981]
(independently) establish necessary and sufficient conditions for
monotone $C^1$ quadratic splines through the insertion of additional
knots and Fritsch [1982] establishes simplified sufficient conditions
for monotone $C^1$ cubic splines that require no knot insertion.
Gregory and Delbourgo [1983] prove necessary and sufficient conditions
for a closed form solution to monotone $C^1$ quadratic rational spline
interpolation with nonlinear boundary equations and an iterative
approach. Delbourgo and Gregory [1993] extend that same work to
achieve sufficient conditions for $C^2$ continuity with a cubic
rational spline given additional tension parameters defined by
users. Huynh [1993] similarly arrives at monotone $C^1$ cubic spline
interpolants and several necessary nonlinear boundary conditions on
monotonicity.

Continuing towards monotone $C^2$ spline interpolation, Fiorot and
Tabka [1991] prove a simple method for determining the existence of a
monotone $C^2$ cubic interpolating spline, but note that functions of
this kind do not always exist for arbitrary convex monotone data sets.
Pruess [1993] proposes sufficient conditions for a cubic $C^2$ shape
preserving spline method while acknowledging that quintic splines are
necessary for generally monotone $C^2$ spline interpolants that do not
insert new knots. Similarly, Manni [1997] constructs sufficient
conditions for a monotone $C^2$ cubic spline by adding two additional
knots per interval, then Cravery and Manni [2003] extend that
methodology to arrive at monotone $C^3$ interpolating splines by
progressively increasing tension through Bezier control
polygons. Constantini [1997] provides a method for monotone $C^2$
quintic spline interpolation based on sufficient and necessary
boundary value conditions, but the necessary conditions are not
sharp. Dougherty, Edelman, and Hyman [1989] construct monotone $C^2$
Hermite splines and in commentary recommend against the direct
optimization for smoothness parameters like global $L^2$, rather
promoting problem specific definitions of geometric niceness. Both
Wang and Tan [2004] as well as Yao and Nelson [2018] propose
sufficient conditions for $C^2$ quartic splines, while Piah and
Unsworth [2011] improve upon those sufficient conditions for $C^2$
quartic rational spline interpolation. Abbas, Majid, and Ali [2012]
formulate a set of sufficient conditions for monotone $C^2$ cubic
rational splines and similar work is extended to $C^2$ quartic
rationals (and splines with arbitrary smoothness) by Zhu and Han
[2015]. Ulrich and Watson [1994] arrived at sufficient and sharp
necessary conditions for monotone $C^2$ quintic splines, as also did
He{\ss} and Schmidt [1994] (independently). In addition, a statistical
method for bootstrapping the construction of an arbitrarily smooth
monotone fit exists [Leitenstorfer and Tutz 2006], but this is enabled
by sufficient conditions only.

The currently available peer reviewed and published software for
monotone piecewise polynomial interpolation is severely reduced in
comparison with the number of published approaches mentioned
above. This is partially indicative of how difficult it is to properly
handle the numerical conditions that arise when constructing and
evaluating precise shape preserving splines, as well as how difficult
it is to create code that correctly conforms with the theoretical
expectations. The sufficient $C^1$ cubic spline method PCHIP of
Fritsch and Carlson [1980] is available through the SciPy Python
package ($>1$ billion downloads), and the $C^1$ quadratic method of
Schumaker [1983] is available as an R package (300 thousand
downloads). The piecewise quadratic $C^1$ method of McAllister and
Roulier [1981] is available in FORTRAN (1000 downloads) and the
sufficient $C^2$ quintic method of Costantini [1997] BVSPIS in FORTRAN
as well (500 downloads). Based on publicly available download records,
it is assumed that the code by Fritsch [1982] for monotone $C^1$ cubic
spline interpolation is the predominant code for constructing monotone
interpolants at present.

The theory for sufficient and {\it sharp} necessary bounds for
monotone quintic interpolation has been provided [Ulrich and Watson
  1994; He{\ss} and Schmidt 1994] and that theory was recently
utilized in a proposed algorithm [Lux 2020] for monotone quintic
spline construction, however no published mathematical software exists
for the quintic case based on sharp monotonicity conditions. The
software presented here represents the first published software
package for producing $C^2$ shape preserving splines based on sharp
monotonicity conditions. This work improves upon the algorithm
presented by Lux et al. [2020] by refactoring computations for
improved numerical stability, estimating minimum magnitude second
derivatives at breakpoints with a quadratic facet model, and using a
binary search to reduce the magnitude of the modifications made to
initial derivative estimates when constructing a monotone spline
interpolant.

@@

This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
sharp necessary and sufficient conditions in Ulrich and Watson [1994]
for the construction of monotone quintic spline interpolants of
monotone data. Precisely, the problem is, given a strictly increasing
sequence $X_1<X_2< \cdots <X_n$ of breakpoints with corresponding
monotone increasing function values $Y_1\le Y_2\le \cdots \le Y_n$,
find a $C^2$ monotone increasing quintic spline $Q(x)$ with the same
breakpoints satisfying $Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI}
actually does something slightly more general, producing $Q(x)$ that
is monotone increasing (decreasing) wherever the data is monotone
increasing (decreasing).)

The remainder of this paper is structured as follows: Section 2 provides
the algorithms for constructing a $C^2$ monotone quintic spline interpolant
to monotone data, Section 3 outlines the method of spline representation
($B$-spline basis) and evaluation, Section 4 analyzes the complexity and
sensitivity of the algorithms in {\tt MQSI}, and Section 5 presents 
timing data and some graphs of constructed interpolants.

>>>>

Many domains of science rely on smooth approximations to real-valued
functions over a closed interval. Piecewise polynomial functions
(splines) provide the smooth approximations for animation in graphics
[Herman and Oftedal\ 2015; Quint 2003], aesthetic structural support in
architecture [Brennan 2020], efficient aerodynamic surfaces in
automotive and aerospace engineering [Brennan 2020], prolonged
effective operation of electric motors [Berglund et\ al.\ 2009], and
accurate nonparametric approximations in statistics [Knott
2012]. While polynomial interpolants and regressors apply broadly,
splines are often a good choice because they can approximate globally
complex functions while minimizing the local complexity of an
approximation.

It is often the case that the true underlying function or phenomenon
being modeled has known properties like convexity, positivity,
various levels of continuity, or monotonicity. Given a reasonable
amount of data, it quickly becomes difficult to achieve desirable
properties in a single polynomial function. In general, the
maintenance of function properties through interpolation/regression is
referred to as {\it shape preserving} [Fritsch and Carlson 1980;
Gregory 1985]. The specific properties the present algorithm will
preserve in approximations are monotonicity and $C^2$ continuity. 
\beginred
Notably this work does not consider convexity preservation, as Mulansky
and Neamtu [1991] prove that there exists strictly convex data sets which
do not allow convex interpolation with $C^2$ splines of fixed degree.
\endred
In addition to previously mentioned applications, \red{being $C^2$ and monotonicity preserving is}
crucially important in statistics to the approximation of a cumulative
distribution function and subsequently the effective generation of
random numbers from a specified distribution [Ramsay 1988].  A spline
function with these properties could approximate a cumulative
distribution function to a high level of accuracy with relatively few
intervals. A twice continuously differentiable approximation to a
cumulative distribution function (CDF) would produce a corresponding
probability density function (PDF) that is continuously
differentiable, which is desirable.

\beginred

The existing research in shape preserving interpolatory splines is
rich and filled with many approaches for many different
applications. In the context of this work, the unique quality of shape
preserving spline algorithms can be observed through: (1) the type of
spline (rational or Hermite), (2) the polynomial order and level of
continuity, and (3) the sharpness, sufficiency, and necessity of the
conditions used to establish shape preserving properties. While the
choice of spline representation is less important, achieving higher
levels of continuity is often more difficult than lower levels of
continuity. Furthermore, all referenced algorithms establish
sufficient conditions for shape preservation, but it is less common
and much more difficult to provide {\it sharp and necessary}
conditions for monotonicity. The following two paragraphs highlight a
collection of research related to the construction of $C^1$ and $C^2$
shape preserving splines. From this foundation, the value and utility
of the code presented here can be more readily appreciated.

DeVore [1977] proves that monotone spline interpolants have improved
accuracy over their nonmonotone counterparts when approximating
monotone functions, while Costantini [1986] proves that monotone
spline interpolants to data of fixed continuity and order exist in
general. This groundwork establishes the potential for monotone spline
interpolation. Schumaker [1983] and McAllister and Roulier [1981a]
(independently) establish necessary and sufficient conditions for
monotone $C^1$ quadratic splines through the insertion of additional
knots and Fritsch [1982] establishes simplified sufficient conditions
for monotone $C^1$ cubic splines that require no knot insertion.
Gregory and Delbourgo [1982] prove necessary and sufficient conditions
for a closed form solution to monotone $C^1$ quadratic rational spline
interpolation with nonlinear boundary equations and an iterative
approach. Delbourgo and Gregory [1983] extend that same work to
achieve sufficient conditions for $C^2$ continuity with a cubic
rational spline given additional tension parameters defined by
users. Huynh [1993] similarly arrives at monotone $C^1$ cubic spline
interpolants and several necessary nonlinear boundary conditions on
monotonicity.

Continuing towards monotone $C^2$ spline interpolation, Fiorot and
Tabka [1991] prove a simple method for determining the existence of a
monotone $C^2$ cubic interpolating spline, but note that functions of
this kind do not always exist for arbitrary convex monotone data sets.
Pruess [1993] proposes sufficient conditions for a cubic $C^2$ shape
preserving spline method while acknowledging that quintic splines are
generally necessary for monotone $C^2$ spline interpolants that do not
insert new knots. Similarly, Manni and Sablonni\`ere [1997] construct
sufficient conditions for a monotone $C^2$ cubic spline by adding two
additional knots per interval, then Cravero and Manni [2003] extend
that methodology to arrive at monotone $C^3$ interpolating splines by
progressively increasing tension through Bezier control
polygons. Costantini [1997] provides a method for monotone $C^2$
quintic spline interpolation based on sufficient and necessary
boundary value conditions, but the necessary conditions are not
sharp. Dougherty, et\ al.\ [1989] construct monotone $C^2$ Hermite
splines and in commentary recommend against the direct optimization
for smoothness parameters like global $L^2$, rather promoting problem
specific definitions of geometric niceness. Both Wang and Tan [2004]
as well as Yao and Nelson [2018] propose sufficient conditions for
$C^2$ quartic splines, while Piah and Unsworth [2011] improve upon
those sufficient conditions for $C^2$ quartic rational spline
interpolation. Abbas, et\ al.\ [2012] formulate a set of sufficient
conditions for monotone $C^2$ cubic rational splines and similar work
is extended to $C^2$ quartic rationals (and splines with arbitrary
smoothness) by Zhu and Han [2015a; 2015b]. Ulrich and Watson [1990]
arrive at sharp necessary and sufficient conditions for monotone $C^2$
quintic splines, while He{\ss} and Schmidt [1994] produce similar
sufficient conditions for $C^2$ quintic splines. In addition, a method
for constructing an arbitrarily smooth monotone fit exists
[Leitenstorfer and Tutz 2006] as well as a method for arbitrarily
smooth shape preservation [Han 2018], but these methods are enabled by
sufficient conditions only.

The currently available peer reviewed and published software for
monotone piecewise polynomial interpolation is severely limited in
comparison with the number of published approaches mentioned
above. This is partially indicative of how difficult it is to properly
handle the numerical conditions that arise when constructing and
evaluating precise shape preserving splines, as well as how difficult
it is to create code that correctly conforms with the theoretical
expectations. The sufficient $C^1$ cubic spline method PCHIP of
Fritsch and Carlson [1980] is available through the SciPy Python
package ($>$1 billion downloads), and the $C^1$ piecewise quadratic
method of Schumaker [1983] is available as an R package (300,000
downloads). The $C^1$ piecewise quadratic method of McAllister and
Roulier [1981a;\ 1981b] is available in FORTRAN (1000 downloads) and
the sufficient $C^2$ piecewise quintic method of Costantini [1997a;
  1997b] BVSPIS in FORTRAN as well (500 downloads). Based on publicly
available download records, it is assumed that the code by Fritsch
[1982] for monotone $C^1$ cubic spline interpolation is the
predominant code for constructing monotone interpolants at present.

\endred

The theory for \red{{\it sharp}} necessary and sufficient bounds for
monotone $C^2$ quintic interpolation has been provided [Ulrich and
  Watson 1990; 1994] and that theory was recently utilized in a
proposed algorithm [Lux et al. 2020; Lux 2020] for monotone $C^2$
quintic spline construction, however no published mathematical
software exists for the quintic case \red{based on sharp monotonicity
conditions}. The software presented here represents the first published
software package for producing $C^2$ shape preserving splines
\red{based on sharp monotonicity conditions}. This work improves upon
the algorithm presented by Lux et\ al.\ [2020] by refactoring
computations for improved numerical stability, estimating minimum
magnitude second derivatives at breakpoints with a quadratic facet
model, and using a binary search to reduce the magnitude of the
modifications made to initial derivative estimates when constructing a
monotone spline interpolant.

\beginred

@@

\endred

This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
sharp necessary and sufficient conditions in Ulrich and Watson [1994]
for the construction of monotone $C^2$ quintic spline interpolants of
monotone data. Precisely, the problem is, given a strictly increasing
sequence $X_1<X_2< \cdots <X_n$ of breakpoints with corresponding
monotone increasing function values $Y_1\le Y_2\le \cdots \le Y_n$,
find a $C^2$ monotone increasing quintic spline $Q(x)$ with the same
breakpoints satisfying $Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI}
actually does something slightly more general, producing $Q(x)$ that
is monotone increasing (decreasing) wherever the data is monotone
increasing (decreasing).)

The remainder of this paper is structured as follows: Section 2
provides the algorithms for constructing a monotone $C^2$ quintic
spline interpolant to monotone data, Section 3 outlines the method of
spline representation ($B$-spline basis) and evaluation, Section 4
analyzes the complexity and sensitivity of the algorithms in {\tt
  MQSI}, and Section 5 presents timing data and some graphs of
constructed interpolants \red{as well as a visual comparison with
  existing monotone spline packages}.

### 2022-03-19 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
All testing code and last comments implemented.

For constructing a quadratic interpolant in $x$ over the interval
$[L,R]$, the Chebyshev basis $1$, $z$, $2z^2-1$ is used, where
$z={x-(L+R)/2 \over (R-L)/2}$. The estimated derivative values by the
quadratic facet model are not guaranteed to produce monotone quintic
polynomial segments. Ulrich and Watson [1994] established tight
constraints on the monotonicity of a quintic polynomial piece, while
deferring to He{\ss} and Schmidt [1994] for a relevant simplified
case. The following algorithm implements a sharp check for
monotonicity by considering the nondecreasing case. The nonincreasing
case is handled similarly.

>>>>

For constructing a quadratic interpolant in $x$ over the interval
$[L,R]$, the Chebyshev basis $1$, $z$, $2z^2-1$ is used, where
$z={x-(L+R)/2 \over (R-L)/2}$. The estimated derivative values by the
quadratic facet model are not guaranteed to produce monotone quintic
polynomial segments. Ulrich and Watson [1990] established tight
constraints on the monotonicity of a quintic polynomial piece and
He{\ss} and Schmidt [1994] gave simpler conditions for a special case.
The following algorithm implements a sharp check for monotonicity by
considering the nondecreasing case. The nonincreasing case is handled
similarly.
